### 2015-09-28 chapters/orthography_profiles.tex
added some TODOs
minor correction 4.1

Given these goals, Unicode locale descriptions (see Section~\ref{terminology})
might seem like the ideal orthography profiles. However, there are various
practical obstacles preventing the use of such locale descriptions in the daily
linguistic practice, namely: (i) the XML-structure is too verbose to easily and
quickly produce or correct manually, (ii) locale descriptions are designed for a
wide scope on information (like date formats or names of weekdays) most of which
is not applicable for documenting writing systems, and (iii) most crucially,
even if someone made the effort to produce a technically correct locale
description for a specific source at hand, then it is nigh impossible to deploy
the description. This is because a locale description has to be submitted to and
accepted by the Unicode Common Locale Data Repository. The repository is
(rightly so) not interested in descriptions that only apply to a limited set of
sources (e.g.~only a single specific dictionary).

The major challenge then is developing an infrastructure to identify the
elements that are individual graphemes in a source, specifically for the
enormous variety of sources using some kind of alphabetic writing system.
Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a
variety of writing systems that range from their own idiosyncratic
transcriptions to already well-established practical or longstanding
orthographies. Although the IPA is one practical choice as a sound-based
normalization for writing systems (which can act as an interlingual pivot to
attain interoperability across writing systems), graphemes in each writing
system must also be identified and standardized if interoperability across
different sources is to be achieved. In most cases, this amounts to more than
simply mapping a grapheme to an IPA segment because graphemes must first be
identified in context (e.g.~is the sequence one sound or two sounds or both?)
and strings must be tokenized, which may include taking orthographic rules into
account (e.g.~between vowels is /n/ and after a vowel but before a consonant is
a nasalized vowel /ṽ/). In our experience, data from each source must be
individually tokenized into graphemes so that its orthographic structure is
identified and its contents can be extracted. To extract data for analysis, a
source-by-source approach is required before an orthography profile can be
created. For example, almost each available lexicon on the world's languages is
idiosyncratic in its orthography and thus requires lexicon-specific approaches
to identify graphemes in the writing system and to map graphemes to phonemes, if
desired.

Thus, our key proposal for the characterization of a writing system is to use a
grapheme tokenization as an inter-orthographic pivot. Basically, any source
document is tokenized by graphemes, and only then a mapping to IPA (or any other
orthographic conversion) is performed. An orthography profile then is a
description of the units and rules that are needed to adequately model a
graphemic tokenization for a language variety as described in a particular
source document. An orthography profile summarizes the Unicode (tailored)
graphemes and orthographic rules used to write a language (the details of the
structure and assumptions of such a profile will be presented in the next
section).

@@

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item code points (10 text elements): t s ʰ o ̃ ̰ ´ s h i 
	\item grapheme clusters (7 text elements): t s ʰ ṍ̰ s h i 
	\item tailored grapheme clusters (4 text elements): tsʰ ṍ̰ sh i 
\end{enumerate}

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
(using NFD normalization), delimited here by space. Unicode tokenization is
required because sequences of code points can differ in their visual and logical
orders. For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> +
<̃> or + <̃> + <̰>. Although these two variants are visually homoglyphs,
computationally they are different. Unicode normalization should be applied to
this string to reorder the code points into a canonical order, allowing the data
to be treated canonically equivalently for search and comparison. In (2), the
Unicode code points have been logically normalized and visually organized into
grapheme clusters, as specified by the Unicode Standard. The combining character
sequence <õ̰> is normalized and visually grouped together. Note that, the
MODIFIER LETTER SMALL H at \uni{02B0}, is not grouped with. This is because it
belongs to Spacing Modifier Letters category in the Unicode Standard. These
characters are underspecified for the direction in which they modify a host
character. For example, can indicate either pre- or post-aspiration (whereas the
nasalization or creaky diacritic is defined in the Unicode Standard to apply to
a specified base character). Finally, to arrive at the graphemic tokenization in
(3), tailored grapheme clusters are needed (as for example specified in an
orthography profile). For example, this orthography profile would specify that
the sequence of characters, and form a single grapheme, and that and form a
grapheme. The orthography profile could also specify orthographic rules,
e.g.~when tokenization graphemes, in say English, the in the forms and should be
treated as distinct sequences depending on their contexts.

>>>>

Given these goals, Unicode locale descriptions (see Section~\ref{terminology})
might seem like the ideal orthography profiles. However, there are various
practical obstacles preventing the use of such locale descriptions in the daily
linguistic practice, namely: (i) the XML-structure is too verbose to easily and
quickly produce or correct manually, (ii) locale descriptions are designed for a
wide scope on information (like date formats or names of weekdays) most of which
is not applicable for documenting writing systems, and (iii) most crucially,
even if someone made the effort to produce a technically correct locale
description for a specific source at hand, then it is nigh impossible to deploy
the description. This is because a locale description has to be submitted to and
accepted by the Unicode Common Locale Data Repository. The repository is
(rightly so) not interested in descriptions that only apply to a limited set of
sources (e.g.~descriptions for only a single dictionary).

The major challenge then is developing an infrastructure to identify the
elements that are individual graphemes in a source, specifically for the
enormous variety of sources using some kind of alphabetic writing system.
Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a
variety of writing systems that range from their own idiosyncratic
transcriptions to already well-established practical or longstanding
orthographies. Although the IPA is one practical choice as a sound-based
normalization for writing systems (which can act as an interlingual pivot to
attain interoperability across writing systems), graphemes in each writing
system must also be identified and standardized if interoperability across
different sources is to be achieved. In most cases, this amounts to more than
simply mapping a grapheme to an IPA segment because graphemes must first be
identified in context (e.g.~is the sequence one sound or two sounds or both?)
and strings must be tokenized, which may include taking orthographic rules into
account (e.g.~between vowels is /n/ and after a vowel but before a consonant is
a nasalized vowel /ṽ/).

In our experience, data from each source must be
individually tokenized into graphemes so that its orthographic structure can be 
identified and its contents can be extracted. To extract data for analysis, a
source-by-source approach is required before an orthography profile can be
created. For example, almost each available lexicon on the world's languages is
idiosyncratic in its orthography and thus requires lexicon-specific approaches
to identify graphemes in the writing system and to map graphemes to phonemes, if
desired.

Our key proposal for the characterization of a writing system is to use a
grapheme tokenization as an inter-orthographic pivot. Basically, any source
document is tokenized by graphemes, and only then a mapping to IPA (or any other
orthographic transliteration) is performed. An \textsc{orthography profile} then is a
description of the units and rules that are needed to adequately model a
graphemic tokenization for a language variety as described in a particular
source document. An orthography profile summarizes the Unicode (tailored)
graphemes and orthographic rules used to write a language (the details of the
structure and assumptions of such a profile will be presented in the next
section).

% TODO: add tsoshi figure here?

@@

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item code points (10 text elements): t s ʰ o \dia{0303} \dia{0330} \dia{0301} s h i 
	\item grapheme clusters (7 text elements): t s ʰ ṍ̰ s h i 
	\item tailored grapheme clusters (4 text elements): tsʰ ṍ̰ sh i 
\end{enumerate}

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
(using NFD normalization), delimited here by space. Unicode tokenization is
required because sequences of code points can differ in their visual and logical
orders. For example, <õ̰> is ambiguous to whether it is the sequence of <o> +
<\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
computationally they are different. Unicode normalization should be applied to
this string to reorder the code points into a canonical order, allowing the data
to be treated canonically equivalently for search and comparison.

In (2), the
Unicode code points have been logically normalized and visually organized into
grapheme clusters, as specified by the Unicode Standard. The combining character
sequence <õ̰> is normalized and visually grouped together. Note that, the
\textsc{modifier letter small h} at \uni{02B0}, is not grouped with any 
other character. This is because it
belongs to Spacing Modifier Letters category in the Unicode Standard. These
characters are underspecified for the direction in which they modify a host
character. For example, can indicate either pre- or post-aspiration (whereas the
nasalization or creaky diacritic is defined in the Unicode Standard to apply to
a specified base character).

Finally, to arrive at the graphemic tokenization in
(3), tailored grapheme clusters are needed (as for example specified in an
orthography profile). For example, this orthography profile would specify that
the sequence of characters, and form a single grapheme, and that and form a
grapheme. The orthography profile could also specify orthographic rules,
e.g.~when tokenization graphemes, in say English, the in the forms and should be
treated as distinct sequences depending on their contexts.
